\begin{vdm}{\small\sf class} $World$\index{World|ModDef}
\par
\kInstanceVarDef
\parlinebr
\begin{insvar}\kw{public}\kw{ }\kw{static}
\assdef{env}{\Opt{Environment\index{Environment|TypeOcc}}}[{\Nil }]
\end{insvar}
\begin{insvar}\kw{public}\kw{ }\kw{static}
\assdef{timerRef}{Timer\index{Timer|TypeOcc}}[{\fnapply{\new{Timer}}{}}]
\end{insvar}
\par
\kOperations
\kw{\kw{public}}\begin{op}[e]{World}\index{World|FuncDef}%
\signature{() \Oto World\index{World|TypeOcc}}
\parms{}
\annlab[o]{World`World}
\begin{blockstmt}
\ass{env}{\fnapply{\new{Environment}}{\Dquote scenario.txt \Dquote }} ; \\
$env$.\call{addSensor}{FighterAircraft`sensor0} ; \\
$env$.\call{addSensor}{FighterAircraft`sensor1} ; \\
$env$.\call{addSensor}{FighterAircraft`sensor2} ; \\
$env$.\call{addSensor}{FighterAircraft`sensor3} ; \\
FighterAircraft`controller0.\call{addDispenser}{FighterAircraft`dispenser0} ; \\
FighterAircraft`controller0.\call{addDispenser}{FighterAircraft`dispenser1} ; \\
FighterAircraft`controller0.\call{addDispenser}{FighterAircraft`dispenser2} ; \\
FighterAircraft`controller0.\call{addDispenser}{FighterAircraft`dispenser3} ; \\
FighterAircraft`detector.\call{addController}{FighterAircraft`controller0} ; \\
FighterAircraft`controller1.\call{addDispenser}{FighterAircraft`dispenser4} ; \\
FighterAircraft`controller1.\call{addDispenser}{FighterAircraft`dispenser5} ; \\
FighterAircraft`controller1.\call{addDispenser}{FighterAircraft`dispenser6} ; \\
FighterAircraft`controller1.\call{addDispenser}{FighterAircraft`dispenser7} ; \\
FighterAircraft`detector.\call{addController}{FighterAircraft`controller1} ; \\
FighterAircraft`controller2.\call{addDispenser}{FighterAircraft`dispenser8} ; \\
FighterAircraft`controller2.\call{addDispenser}{FighterAircraft`dispenser9} ; \\
FighterAircraft`controller2.\call{addDispenser}{FighterAircraft`dispenser10} ; \\
FighterAircraft`controller2.\call{addDispenser}{FighterAircraft`dispenser11} ; \\
FighterAircraft`detector.\call{addController}{FighterAircraft`controller2}
\end{blockstmt};
\end{op}
\kw{\kw{public}}\begin{op}[e]{Run}\index{Run|FuncDef}%
\signature{() \Oto ()}
\parms{}
\annlab[o]{World`Run}
$env$.\call{Run}{}
\end{op}
{\small\sf end} {\it World}

\end{vdm}

















































\begin{tabular}{p{25mm}l}
{\bf Test Suite :} & vdm.tc \\ 
{\bf Class :} & World \\ 
\end{tabular}

\begin{longtable}{|l|r|r|}\hline
{\bf World`World} & {\bf \#Calls} & {\bf Coverage} \kill
{\bf Name} & {\bf \#Calls} & {\bf Coverage} \\ \hline\hline
\endhead
World`Run & 1 & $\surd$ \\ \hline
World`World & 1 & $\surd$ \\ \hline
\hline
{\bf Total Coverage} & & {\bf 100\%} \\ \hline
\end{longtable}


